---
title: "CLV Case Study"
author: "Wei Miao"
institute: UCL School of Management
df-print: kable
colorlinks: true
code-line-numbers: true
format:
  html: 
    toc: true
    number-sections: true
    page-layout: full
    toc-depth: 2
    code-line-numbers: true
    code-copy: hover
knitr:
  opts_chunk:
    echo: true
    warning: true
    message: true
    error: true
execute: 
  freeze: auto
---

# Situation Analysis

Since i-basket is a US company. Let's use Marks and Spencer as a UK example for conducting 5C analysis for a grocery retailer. You can follow similar templates to conduct situation analysis for the first assignment.

**1. Company**

-   **Products**: M&S provides a wide range of products, including groceries, clothing, cafe, and home products.

-   **Value Proposition**: Offers high-quality, sustainable products and is known for its strong ethical standards.

-   **Market Position**: Holds a steady market share, but faces challenges in some segments such as clothing and home segments.

**2. Customers**

-   **Demographics**: Appeals to a broad demographic but is notably popular among middle to upper-middle-class consumers.

-   **Customer Needs**: Customers seek quality groceries, fashionable clothing, and home goods.

-   **Customer Journey**: Engaging through both online and offline channels, M&S appeals to both in-store and e-commerce shoppers.

**3. Competitors**

-   **Direct Competitors**: Tesco, Sainsbury's, and other UK retail giants.

-   **Indirect Competitors**: Online retailers and fast-fashion brands.

-   **Unique Selling Propositions (USPs)**: Competitors may offer lower prices (Lidl, Aldi) or broader product ranges (Tesco).

**4. Collaborators**

-   **Suppliers**: Partnerships with farmers and manufacturers.

-   **Distributors**: Works through its own retail network and an online platform.

-   **Joint Ventures**: Various collaborations, like with Ocado for online grocery retailing.

**5. Climate**

-   **Economic Factors**: Influenced by economic health and consumer spending patterns.

-   **Legal & Political Factors**: Brexit and changing trade and labor laws impact operations.

-   **Social & Cultural Trends**: Growing preference for online shopping and sustainable practices.

-   **Technological Innovations**: The integration of technology in retail experience and e-commerce. Lack direct online channel, may need to develop it's own grocery shopping online and mobile channels to keep up with competitors.

-   **Environmental Factors**: Adopting and promoting sustainable and ethical practices.

# Compute Customer Lifetime Value

## Step 1: Determine time unit of analysis

-   Time unit of analysis
    -   *\[...\] Because the typical customer contract is renewed annually, all relevant variables will be measured on a yearly basis.*

        -   When should we use monthly analysis? When there is strong within-year seasonality of customer purchases

## Step 2: Determine number of years

-   $N$: the number of years over which the customer relationship is assessed
    -   *\[...\] i-basket typically used a five-year horizon for revenue and profitability calculations*

```{r}
N <- 5
```

## Step 3: Compute profit margin for each period

$CF = M - c$**: gross profit each year, which is the profit from sales M minus marketing costs c**

-   *most customers paid the \$99 annual membership fee*

```{r}
membership <- 99
```

-   *40 times each year; each time \$100; with profit margin 7% (COGS 93%)*

```{r}
n_visit <- 40
revenue_each_visit <- 100
```

-   *profit margin 7% (COGS 93%)*

```{r}
profit_margin <- 0.07
M <- membership + revenue_each_visit * n_visit * profit_margin
```

-   *variable delivery costs each order*

```{r}
deliverycost_each_visit <- 5 + revenue_each_visit * 0.035
c <- deliverycost_each_visit * n_visit
```

-   the annual CF from customers CF

```{r}

CF <- M - c
profit_seq <- rep(CF,N)
profit_seq

```

## Step 4: Compute sequence of retention rate

4.  $r$: retention rate

-   *\[...\] i-basket examined the proportion of customers who renewed their membership from one year to the next (using an average across three years). This proportion was 0.7*

```{r}
# retention_rate is the probability of customer staying with us after 1 year
retention_rate <- 0.7

# create a geometric sequence of accumulative retention rate for N years  
retention_seq <- retention_rate ^ (seq(1,N) - 1)
```

## Step 5: Compute sequence of discount factors

5.  $k$: the interest rate

-   *\[...\] A yearly discount rate of 10%*

```{r}
discount_rate <- 0.1
discount_factor <- 1/(1+discount_rate)
discount_factor_seq <- discount_factor ^ (seq(1,N))
discount_factor_seq
```

-   *\[...\] The team decided to take a conservative approach whereby all profits are booked at the end of year*.
    -   All profits earned per customer in year 1 need to be discounted once, the profits earned in year 2 need to be discounted twice, and so on

## Step 6: Compute customer acquisition costs

6.  CAC = total costs for customer ad clicks + total costs of \$15 promo + total costs of free deliveries

    **CAC Part I: Costs of paid search ads to get 1 new member.**

-   *\[...\] 20% of those that signed up for the free trial ended up becoming members*

-   *\[...\] a fifth of those who clicked on an ad were willing to give the service a try*

```{r}
# clicker_to_trier_rate is the % of trier customers from clickers
clicker_to_trier_rate <- 0.2

# trier_to_member_rate is the % of a new member from triers
trier_to_member_rate <-  0.2
```

-   How many customers need to click the ad to get 1 new customer?

```{r}
n_clickers_for_1newmember <- (1/clicker_to_trier_rate) * (1/trier_to_member_rate)

n_clickers_for_1newmember
```

-   Total costs for customer clicks

```{r}
total_cost_clicks <- 0.4 * n_clickers_for_1newmember
total_cost_clicks
```

**CAC Part II: total costs of \$15 promo for first order each trier customer**

-   What is the total promo cost for these trier customers' first order?

```{r}
promo_first_order_each_trier <- 15

total_cost_promo <- promo_first_order_each_trier * # promotion amount
  (1 - profit_margin) *    # profit rate
  (1/trier_to_member_rate) # num of triers

total_cost_promo
```

**CAC Part III: total costs from selling groceries for each trier**

-   2 visits for each trier, the profits from the 2 visits are

```{r}

profit_each_trier <-  revenue_each_visit * profit_margin * 2

profit_each_trier
```

-   The 2 visits are free of delivery charges, which are costs to i-basket

```{r}
deliverycost_1st <- 5 + (revenue_each_visit + promo_first_order_each_trier) * 0.035
deliverycost_2nd <- 5 + revenue_each_visit * 0.035
deliverycost_each_trier <- deliverycost_1st + deliverycost_2nd
deliverycost_each_trier
```

-   Net costs for each trier from the 2 visits

```{r}

netcost_each_trier <- deliverycost_each_trier - profit_each_trier

netcost_each_trier
```

-   Total net profits from all 25 triers

```{r}
totalcosts_from_all_triers <- netcost_each_trier * (1/trier_to_member_rate)

totalcosts_from_all_triers
```

-   CAC = total costs for customer ad clicks + total costs of \$15 promo + total costs of free deliveries

```{r}
CAC <- total_cost_clicks + total_cost_promo + totalcosts_from_all_triers

CAC
```

## Step 7: Compute CLV

7.  Compute the CLV based on the CLV formula (Table A)

-   Revenues, variables costs, and profit for the next 5 years

```{r}
profit_seq
```

-   Apply retention rate

```{r}
profit_seq_after_churn <- profit_seq * retention_seq
profit_seq_after_churn
```

-   Apply discount factor

```{r}
profit_seq_after_churn_discount<- profit_seq_after_churn * discount_factor_seq

profit_seq_after_churn_discount
```

-   Compute CLV by summing up future expected profits

```{r}
sum(profit_seq_after_churn_discount)

sum(profit_seq_after_churn_discount) - CAC
```

# Use CLV to Guide Marketing Decisions

We can use CLV to guide our marketing decisions. Remember that any marketing activity comes with a cost and a benefit.

## User Defined Function

To facilitate the process, let's first see how to use **a user defined function** to compute CLV for any new scenario with ease.

```{r}

computeCLV <- function(N = 5,
                       membership = 99,
                       n_visit = 40,
                       revenue_each_visit = 100,
                       profit_margin = 0.07,
                       retention_rate = 0.7,
                       discount_rate = 0.1,
                       clicker_to_trier_rate = 0.2,
                       trier_to_member_rate = 0.2,
                       promo_first_order_each_trier = 15) {
  
  # Step 3
  
  ## compute M
  M <- membership + revenue_each_visit * n_visit * profit_margin
  
  ## compute c
  deliverycost_each_visit <- 5 + revenue_each_visit * 0.035
  
  c <- deliverycost_each_visit * n_visit
  
  ## compute CF = M - c
  CF <- M - c
  
  ## compute profit sequence
  profit_seq <- rep(CF,N)
  
  # Step 4: compute sequence of retention
  retention_seq <- retention_rate ^ (seq(1,N) - 1)
  
  # Step 5: compute sequence of discount factors
  discount_factor <- 1/(1+discount_rate)
  
  discount_factor_seq <- discount_factor ^ (seq(1,N))
  
  # Step 6: Compute CAC
  
  ## Part I
  n_clickers_for_1newmember <- (1/clicker_to_trier_rate) * (1/trier_to_member_rate)
  
  total_cost_clicks <- 0.4 * n_clickers_for_1newmember
  
  ## Part II
  total_cost_promo <- promo_first_order_each_trier * # promotion amount
    (1 - profit_margin) *    # profit rate
    (1/trier_to_member_rate) # num of triers
  
  ## Part III
  profit_each_trier <-  revenue_each_visit * profit_margin * 2
  
  deliverycost_1st <- 5 + (revenue_each_visit + promo_first_order_each_trier) * 0.035
  deliverycost_2nd <- 5 + revenue_each_visit * 0.035
  deliverycost_each_trier <- deliverycost_1st + deliverycost_2nd
  
  netcost_each_trier <- deliverycost_each_trier - profit_each_trier
  
  totalcosts_from_all_triers <- netcost_each_trier * (1/trier_to_member_rate)
  
  ## Compute CAC
  CAC <- total_cost_clicks + total_cost_promo + totalcosts_from_all_triers
  
  
  # Step 7: Compute CLV
  
  ## apply churn rate to profit sequence
  profit_seq_after_churn <- profit_seq * retention_seq
  
  ## apply discount factor to profit sequence
  profit_seq_after_churn_discount<- profit_seq_after_churn * discount_factor_seq
  
  ## take sum and deduct the CAC
  CLV <- sum(profit_seq_after_churn_discount) - CAC

  return(CLV)
}
```

```{r}
computeCLV()
```

## Use CLV to Guide Marketing Decisions

-   (To guide customer acquisition) What if the company only offers \$5 for first time purchase? This will save some CAC but the cliker-to-trier rate will decrease to 10%. Please compute the new CLV. Should you go ahead with the proposed change?

```{r}
computeCLV(promo_first_order_each_trier = 5,
           clicker_to_trier_rate = 0.1)
```

-   (To guide customer retention) What if the company increases the annual membership fee to \$119? This will increase revenue from memberships but will also make some customers unhappy so their retention rate reduce to 60%. Please compute the new CLV. Should you go ahead with the proposed change? What if the retention rate reduces to 50%?

```{r}
computeCLV(membership = 119 ,
           retention_rate = 0.6)
```

```{r}
computeCLV(membership = 119 ,
           retention_rate = 0.5)
```
