---
author: Dr Wei Miao
date: "`r (lubridate::ymd('20231004')+lubridate::dweeks(4))`"
date-format: long
institute: UCL School of Management
title: "Class 9 Case study: Improve Marketing Efficiency for Tesco Using Supervised Learning"
df-print: kable
colorlinks: true
code-line-numbers: true
format:
  html: 
    toc: true
    number-sections: true
    page-layout: full
    toc-depth: 2
    code-line-numbers: true
    code-copy: hover
  beamer: 
    toc: false
    toc-title: ""
    slide-level: 2
    section-titles: true
    theme: Frankfurt
    colortheme: beaver
    fonttheme: structurebold
    navigation: horizontal
    tbl-colwidths: auto
    fontsize: 9pt
    suppress-bibliography: true
knitr:
  opts_chunk:
    echo: true
    warning: true
    message: true
    error: true
execute: 
  freeze: auto
editor_options: 
  chunk_output_type: inline
bibliography: references.bib
---

# Business Objective

## Background

Tesco is looking to promote its new [private-label](https://en.wikipedia.org/wiki/Private_label) products to existing customers. The marketing analytics team decides to use the conventional mailing marketing strategy so that customer would receive color-printed leaflets via Royal Mails to their doorsteps.

## Cost-Benefit Analyses

-   **Cost**: Each mail costs **£1.5** **to produce** and another **£0.5** **to mail** to the customers.
-   The cost is the marketing offer we send, `cost_per_offer`

```{r}
# cost of sending an offer
cost_per_offer <- 1.5 + 0.5
cost_per_offer
```

::: {.content-visible when-format="beamer"}
## Cost-Benefit Analyses

-   **Benefit**: If customer responds to the offer, the management expects customers to spend £20 on trying the new products, where the COGS is 60%.

-   The benefit is the profit margin if a customer responds, `profit_per_customer`

```{r}
# profit from a responding customer
COGS <- 0.6
profit_per_customer <- 20 * (1 - COGS)
profit_per_customer
```
:::

::: {.content-visible when-format="html"}
-   **Benefit**: If customer responds to the offer, the management expects customers to spend £20 on trying the new products, where the COGS is 60%.

-   The benefit is the profit margin if a customer responds, `profit_per_customer`

```{r}
# profit from a responding customer
COGS <- 0.6
profit_per_customer <- 20 * (1 - COGS)
profit_per_customer
```
:::

## Break-Even Analysis: Break-Even Response Rate

-   In order to break-even, we can calculate the break-even response rate from customers, which is the **minimum response rate** we need of a customer in order not to lose money from sending the marketing offer[^1]

[^1]: The idea break-even is similar to the break-even quantity we learned in Week 1, the minimum incremental quantity we need to sell in order not to lose any money

```{r}
break_even_response <- cost_per_offer/profit_per_customer
break_even_response
```

-   Only if a customer responds to us with at least `r break_even_response*100`% response rate can we recover the costs of making an marketing offer.
-   If we send offers to customers whose expected response rate is lower than `r break_even_response*100`%, we make a loss by expectation.

# Data Analytics

-   **Data collection and cleaning**

    -   Split the data into a training set and a test set

-   **Data analytics**

    -   Train predictive models on the training set

    -   Predict customer response rate on the test set

-   **Business recommendations**

    -   Target customers based on predicted response rate

    -   Compute and compare ROIs for each targeting method: (1) blanket marketing; (2) decision tree; (3) random forest

**Let's work on the Quarto document together!**
