---
title: "CLV Case Study"
author: "Wei Miao"
institute: UCL School of Management
df-print: kable
colorlinks: true
code-line-numbers: true
format:
  html: 
    toc: true
    number-sections: true
    page-layout: full
    toc-depth: 2
    code-line-numbers: true
    code-copy: hover
knitr:
  opts_chunk:
    echo: true
    warning: true
    message: true
    error: true
---

# Step 1: Determine time unit of analysis

-   Time unit of analysis
    -   *\[...\] Because the typical customer contract is renewed annually, all relevant variables will be measured on a yearly basis.*

        -   When should we use monthly analysis? When there is strong within-year seasonality of customer purchases

# Step 2: Determine number of years

-   $N$: the number of years over which the customer relationship is assessed
    -   *\[...\] i-basket typically used a five-year horizon for revenue and profitability calculations*

```{r}
N <- 5
```

# Step 3: Compute profit margin for each period

$CF = M - c$: **gross profit each year**

-   *most customers paid the \$99 annual membership fee*

```{r}
membership <- 99
```

-   *40 times each year; each time \$100*

```{r}
n_visit <- 40
revenue_each_visit <- 100
```

-   *profit margin 7% (COGS 93%)*

```{r}
profit_margin <- 0.07
M <- membership + revenue_each_visit * n_visit * profit_margin
```

-   *variable delivery costs each order*

```{r}
deliverycost_each_visit <- 5 + 100 * 0.035
c <- deliverycost_each_visit * n_visit
```

-   the annual CF from customers CF

```{r}

CF <- M - c
profit_seq <- rep(CF,N)
profit_seq

```

# Step 4: Compute sequence of retention rate

4.  $r$: retention rate

-   *\[...\] i-basket examined the proportion of customers who renewed their membership from one year to the next (using an average across three years). This proportion was 0.7*

```{r}
retention_rate <- 0.7
retention_seq <- retention_rate ^ (seq(1,N) - 1)
```

# Step 5: Compute sequence of discount factors

5.  $k$: the discount rate

-   *\[...\] A yearly discount rate of 10%*

```{r}
discount_rate <- 0.1
discount_factor_seq <- (1 / (1+discount_rate)) ^ (seq(1,N))
discount_factor_seq
```

-   *\[...\] The team decided to take a conservative approach whereby all profits are booked at the end of year*.
    -   All profits earned per customer in year 1 need to be discounted once, the profits earned in year 2 need to be discounted twice, and so on

# Step 6: Compute customer acquisition costs

6.  CAC = total costs for customer ad clicks + total costs of \$15 promo + total costs of free deliveries

1.  **Total costs for customer clicks**

-   *\[...\] 20% of those that signed up for the free trial ended up becoming members*

-   *\[...\] a fifth of those who clicked on an ad were willing to give the service a try*

```{r}
conversion_rate <- 0.2
click_through_rate <- 0.2
```

-   How many customers need to click the ad to get 1 new customer?

```{r}
n_clicks_1newcustomer <- 1/click_through_rate/conversion_rate

```

-   Total costs for customer clicks

```{r}
total_cost_clicks <- 0.4 * n_clicks_1newcustomer
total_cost_clicks
```

(2) **total costs of \$15 promo for first order each trier customer**

-   How many customers need to try the service to get 1 new customer?

```{r}
n_triers <- 1/conversion_rate

```

-   What is the total promo cost for these trier customers' first order?

```{r}
promo_first_order_each_trier <- 15

total_cost_promo <- promo_first_order_each_trier * (1 - profit_margin) * n_triers
total_cost_promo
```

(3) **total costs from free deliveries**

-   Assume two visits, the delivery costs for each visit

```{r}
deliverycost_1st <- 5 + 115 * 0.035
deliverycost_2nd <- 5 + 100 * 0.035
deliverycost_each_trier <- deliverycost_1st + deliverycost_2nd
```

-   We also make a profit each trier

```{r}
profit_each_trier <-  revenue_each_visit * profit_margin * 2
```

-   Net delivery costs each trier

```{r}

NetDeliveryCost_each_trier<- deliverycost_each_trier - profit_each_trier

total_cost_delivery <- NetDeliveryCost_each_trier * n_triers

total_cost_delivery
```

(4) **CAC**

```{r}
CAC <- total_cost_clicks + total_cost_promo + total_cost_delivery
CAC
```

# Step 7: Compute CLV

7.  Compute the CLV based on the CLV formula (Table A)

<!-- -->

(1) Revenues, variables costs, and profit for the next 5 years

```{r}
profit_seq
```

(2) Apply retention rate

```{r}
profit_seq_after_churn <- profit_seq * retention_seq
profit_seq_after_churn
```

(3) Apply discount factor

```{r}
profit_seq_after_churn_discount<- profit_seq_after_churn * discount_factor_seq

profit_seq_after_churn_discount
```

(4) Compute CLV by summing up future expected profits

```{r}
sum(profit_seq_after_churn_discount)
sum(profit_seq_after_churn_discount) - CAC
```
